---
title: 'Age Specific Mortality Rate Trends '
output:
  html_document:
    df_print: paged
---

# Intro

Countries of interest:

* Japan
* UK
* USA
* France
* Canada

Analyses

* e0 for all countries (implied?)
* Mx for age 0, 40, 80, 90
    * Correlation between trends and log-trends
* Mx at all ages 
    * Correlation as heatmap
    * Separated into two periods: 
        * 1975-1990
        * After 1990 (to last common available year?)


# Prereqs 

```{r}
pacman::p_load(
  tidyverse, ggrepel, here,
  corrr, knitr, kableExtra
)

dta_e0 <- read_rds(here("data", "e0.rds"      ))
dta_Mx <- read_rds(here("data", "Mx.rds"      ))

source(here("scripts", "country_definitions.R"))


```

# e0

```{r}
dta_e0 %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y = e0, group = Country, colour = Country, linetype = Country)) + 
  geom_line() + 
  facet_wrap(~sex) +
  labs(
    x = "Year", y = "Life expectancy at birth",
    title = "Life expectancy at birth for selected countries",
    caption = "Source: Human Mortality Database"
  ) + 
  theme_minimal() 

ggsave(here("figures", "00_e0.png"), height = 15, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "00_e0.svg"), height = 15, width = 20, units = "cm", dpi = 300)

```

Annual change in life expectancy 

```{r}
dta_e0 %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  group_by(Country, sex) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  ungroup() %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y = ch_e0)) + 
  geom_point() + stat_smooth() + 
  facet_grid(Country~sex) +
  labs(
    x = "Year", y = "Change in Life expectancy at birth",
    title = "Annual changes in Life expectancy at birth for selected countries",
    subtitle = "Lines show smoothed trend",
    caption = "Source: Human Mortality Database"
  ) +
  geom_hline(yintercept = 0) +
  theme_minimal() 

ggsave(here("figures", "01_ch_e0.png"), height = 20, width = 15, units = "cm", dpi = 300)
ggsave(here("figures", "01_ch_e0.svg"), height = 20, width = 15, units = "cm", dpi = 300)


```



# Mx at specific ages 

```{r}
dta_Mx %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  filter(age %in% c(0, 40, 80, 90)) %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y =  Mx, group = Country, colour = Country, linetype = Country)) +
  geom_line() +
  facet_grid(age ~ sex, scales = "free_y") +
  scale_y_log10() +
  theme_minimal() + 
  labs(
    x = "Year", y = "Probability of death in next 12 months (log scale)",
    title = "Probability of dying in next 12 months by age in years",
    caption = "Source: Human Mortality Database"
  )

ggsave(here("figures", "02_Mx_selected.png"), height = 20, width = 18, units = "cm", dpi = 300)
ggsave(here("figures", "02_Mx_selected.svg"), height = 20, width = 18, units = "cm", dpi = 300)



```

Correlation between log improvement rates by country 

```{r}
corrstretch <- function(x){
  x %>% correlate() %>% stretch()
}

mx_corrs <- 
  dta_Mx %>% 
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
    filter(age %in% c(0, 40, 80, 90)) %>% 
    filter(year >= 1975) %>% 
    filter(sex != "total") %>% 
    mutate(lnmx = log(Mx)) %>%
    select(-Mx) %>% 
    select(Country, sex, age, year, lnmx) %>% 
    spread(age, lnmx) %>% 
    select(-year) %>% 
    group_by(sex, Country) %>% 
    nest() %>% 
    mutate(correlations = map(data, corrstretch)) %>% 
    select(Country, sex, correlations) %>% 
    unnest(correlations) 

mx_corrs  
```

Rearrange for table 

```{r}
corrs_female <- 
  mx_corrs %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "female") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "f_0", "f_40", "f_80", "f_90"))

corrs_male <- 
  mx_corrs %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "male") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "m_0", "m_40", "m_80", "m_90"))

corrs_wide <- full_join(corrs_female, corrs_male, by = c("Country", "age"))
corrs_wide
```

Now to display as a nice table 

```{r}
options(knitr.kable.NA = '')

corrs_wide %>% 
  kable(col.names = c("Country", "Age", "0", "40", "80", "90", "0", "40", "80", "90")) %>% 
  kable_styling(c("bordered")) %>% 
  add_header_above(c(" " = 2, "Female" = 4, "Male" = 4)) %>% 
  collapse_rows(columns = 1)
  
```


Now the same but split by period (1975-1990; 1991-last year)


```{r}
mx_corrs_1975_1990 <- 
  dta_Mx %>% 
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
    filter(age %in% c(0, 40, 80, 90)) %>% 
    filter(between(year, 1975, 1990)) %>% 
    filter(sex != "total") %>% 
    mutate(lnmx = log(Mx)) %>%
    select(-Mx) %>% 
    select(Country, sex, age, year, lnmx) %>% 
    spread(age, lnmx) %>% 
    select(-year) %>% 
    group_by(sex, Country) %>% 
    nest() %>% 
    mutate(correlations = map(data, corrstretch)) %>% 
    select(Country, sex, correlations) %>% 
    unnest(correlations) 

mx_corrs_1975_1990  

```

```{r}
corrs_female_1975_1990 <- 
  mx_corrs_1975_1990 %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "female") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "f_0", "f_40", "f_80", "f_90"))

corrs_male_1975_1990 <- 
  mx_corrs_1975_1990 %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "male") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "m_0", "m_40", "m_80", "m_90"))

corrs_wide_1975_1990 <- full_join(corrs_female_1975_1990, corrs_male_1975_1990, by = c("Country", "age"))
corrs_wide_1975_1990
```



Now to display as a nice table 

```{r}
options(knitr.kable.NA = '')

corrs_wide_1975_1990 %>% 
  kable(col.names = c("Country", "Age", "0", "40", "80", "90", "0", "40", "80", "90")) %>% 
  kable_styling(c("bordered")) %>% 
  add_header_above(c(" " = 2, "Female" = 4, "Male" = 4)) %>% 
  collapse_rows(columns = 1)

```


And for the latter period 


```{r}
mx_corrs_1991_last <- 
  dta_Mx %>% 
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
    filter(age %in% c(0, 40, 80, 90)) %>% 
    filter(year >= 1991) %>% 
    filter(sex != "total") %>% 
    mutate(lnmx = log(Mx)) %>%
    select(-Mx) %>% 
    select(Country, sex, age, year, lnmx) %>% 
    spread(age, lnmx) %>% 
    select(-year) %>% 
    group_by(sex, Country) %>% 
    nest() %>% 
    mutate(correlations = map(data, corrstretch)) %>% 
    select(Country, sex, correlations) %>% 
    unnest(correlations) 

mx_corrs_1991_last

```

```{r}
corrs_female_1991_last <- 
  mx_corrs_1991_last %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "female") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "f_0", "f_40", "f_80", "f_90"))

corrs_male_1991_last <- 
  mx_corrs_1991_last %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "male") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "m_0", "m_40", "m_80", "m_90"))

corrs_wide_1991_last <- full_join(corrs_female_1991_last, corrs_male_1991_last, by = c("Country", "age"))
corrs_wide_1991_last
```



Now to display as a nice table 

```{r}
options(knitr.kable.NA = '')

corrs_wide_1991_last %>% 
  kable(col.names = c("Country", "Age", "0", "40", "80", "90", "0", "40", "80", "90")) %>% 
  kable_styling(c("bordered")) %>% 
  add_header_above(c(" " = 2, "Female" = 4, "Male" = 4)) %>% 
  collapse_rows(columns = 1)

```



# Mx for all ages (heatmap)




```{r}
dta_trnd <- dta_Mx %>% 
  filter(sex != "total") %>%
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
  filter(year >= 1975)  %>% 
  filter(age <= 109) %>% 
  group_by(Country, sex, year, age) %>% 
  summarise(mean_Mx = mean(Mx, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(log_mean_Mx = log(mean_Mx, 10)) %>% 
  group_by(sex, Country) %>% 
  nest()


cors_df <- dta_trnd %>% 
  mutate(cors = map(
    data, 
    function(X) {
      X %>% 
        select(-mean_Mx) %>% 
        spread(age, log_mean_Mx) %>% 
        select(-year) %>% 
        cor()
      }
    )
  ) %>% 
  mutate(cor_df = map(
    cors,
    function(X){
      X %>% 
        as_tibble() %>% 
        mutate(from_age = rownames(X)) %>% 
        gather(key = "to_age", value = "value", -from_age) %>% 
        mutate(from_age = as.numeric(from_age), to_age = as.numeric(to_age))
      }
    )
  ) %>% 
  select(sex, Country, cor_df) %>% 
  unnest()


cors_df %>% 
  filter(from_age <= 100, to_age <= 100) %>% 
  rename(r = value) %>% 
  ggplot(aes(x = from_age, y = to_age, fill = r)) + 
  geom_tile() +
  scale_fill_distiller(palette = "RdBu", limits = c(-1,1), direction = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  coord_equal() + 
  facet_grid(Country ~ sex) +
  labs(x = "", y = "",
       title = "Correlation between % improvement\ntrends at specific ages on x and y axes")

ggsave(here("figures", "heatmap_allyears.png"), width = 18, height = 35, units = "cm", dpi = 300)
ggsave(here("figures", "heatmap_allyears.svg"), width = 18, height = 35, units = "cm", dpi = 300)


```

Now the same for first 1975-1990, then 1991-last year 

```{r}
dta_trnd <- dta_Mx %>% 
  filter(sex != "total") %>%
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
  filter(between(year, 1975, 1990))  %>% 
  filter(age <= 109) %>% 
  group_by(Country, sex, year, age) %>% 
  summarise(mean_Mx = mean(Mx, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(log_mean_Mx = log(mean_Mx, 10)) %>% 
  group_by(sex, Country) %>% 
  nest()


cors_df <- dta_trnd %>% 
  mutate(cors = map(
    data, 
    function(X) {
      X %>% 
        select(-mean_Mx) %>% 
        spread(age, log_mean_Mx) %>% 
        select(-year) %>% 
        cor()
      }
    )
  ) %>% 
  mutate(cor_df = map(
    cors,
    function(X){
      X %>% 
        as_tibble() %>% 
        mutate(from_age = rownames(X)) %>% 
        gather(key = "to_age", value = "value", -from_age) %>% 
        mutate(from_age = as.numeric(from_age), to_age = as.numeric(to_age))
      }
    )
  ) %>% 
  select(sex, Country, cor_df) %>% 
  unnest()


cors_df %>% 
  filter(from_age <= 100, to_age <= 100) %>% 
  rename(r = value) %>% 
  ggplot(aes(x = from_age, y = to_age, fill = r)) + 
  geom_tile() +
  scale_fill_distiller(palette = "RdBu", limits = c(-1,1), direction = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  coord_equal() + 
  facet_grid(Country ~ sex) +
  labs(x = "", y = "",
       title = "Correlation between % improvement\ntrends at specific ages on x and y axes",
       subtitle = "1975-1990")

ggsave(here("figures", "heatmap_1975_1990.png"), width = 18, height = 35, units = "cm", dpi = 300)
ggsave(here("figures", "heatmap_1975_1990.svg"), width = 18, height = 35, units = "cm", dpi = 300)


```


And now 1991 to last available year 

```{r}
dta_trnd <- dta_Mx %>% 
  filter(sex != "total") %>%
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
  filter(year >= 1991)  %>% 
  filter(age <= 109) %>% 
  group_by(Country, sex, year, age) %>% 
  summarise(mean_Mx = mean(Mx, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(log_mean_Mx = log(mean_Mx, 10)) %>% 
  group_by(sex, Country) %>% 
  nest()


cors_df <- dta_trnd %>% 
  mutate(cors = map(
    data, 
    function(X) {
      X %>% 
        select(-mean_Mx) %>% 
        spread(age, log_mean_Mx) %>% 
        select(-year) %>% 
        cor()
      }
    )
  ) %>% 
  mutate(cor_df = map(
    cors,
    function(X){
      X %>% 
        as_tibble() %>% 
        mutate(from_age = rownames(X)) %>% 
        gather(key = "to_age", value = "value", -from_age) %>% 
        mutate(from_age = as.numeric(from_age), to_age = as.numeric(to_age))
      }
    )
  ) %>% 
  select(sex, Country, cor_df) %>% 
  unnest()


cors_df %>% 
  filter(from_age <= 100, to_age <= 100) %>% 
  rename(r = value) %>% 
  ggplot(aes(x = from_age, y = to_age, fill = r)) + 
  geom_tile() +
  scale_fill_distiller(palette = "RdBu", limits = c(-1,1), direction = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  coord_equal() + 
  facet_grid(Country ~ sex) +
  labs(x = "", y = "",
       title = "Correlation between % improvement\ntrends at specific ages on x and y axes",
       subtitle = "1991 onwards")

ggsave(here("figures", "heatmap_1991_last.png"), width = 18, height = 35, units = "cm", dpi = 300)
ggsave(here("figures", "heatmap_1991_last.svg"), width = 18, height = 35, units = "cm", dpi = 300)


```

