---
title: 'Age Specific Mortality Rate Trends '
output:
  html_document:
    df_print: paged
---

# Intro

Countries of interest:

* Japan
* UK
* USA
* France
* Canada

Analyses

* e0 for all countries (implied?)
* Mx for age 0, 40, 80, 90
    * Correlation between trends and log-trends
* Mx at all ages 
    * Correlation as heatmap
    * Separated into two periods: 
        * 1975-1990
        * After 1990 (to last common available year?)


# Prereqs 

```{r}
pacman::p_load(
  tidyverse, ggrepel, here,
  corrr, knitr, kableExtra, openxlsx
)

dta_e0 <- read_rds(here("data", "e0.rds"      ))
dta_Mx <- read_rds(here("data", "Mx.rds"      ))

source(here("scripts", "country_definitions.R"))


```


Write-to-table function 

```{r}
write_to_table <- function(x, location){
  x %>% write_csv(location)
  x
}

```


# e0

```{r}
dta_e0 %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y = e0, group = Country, colour = Country, linetype = Country)) + 
  geom_line() + 
  facet_wrap(~sex) +
  labs(
    x = "Year", y = "Life expectancy at birth",
    title = "Life expectancy at birth for selected countries",
    caption = "Source: Human Mortality Database"
  ) + 
  theme_minimal() 

ggsave(here("figures", "00_e0.png"), height = 15, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "00_e0.svg"), height = 15, width = 20, units = "cm", dpi = 300)

```

Annual change in life expectancy 

```{r}
dta_e0 %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  group_by(Country, sex) %>% 
  arrange(year) %>% 
  mutate(ch_e0 = e0 - lag(e0)) %>% 
  ungroup() %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y = ch_e0)) + 
  geom_point() + stat_smooth() + 
  facet_grid(Country~sex) +
  labs(
    x = "Year", y = "Change in Life expectancy at birth",
    title = "Annual changes in Life expectancy at birth for selected countries",
    subtitle = "Lines show smoothed trend",
    caption = "Source: Human Mortality Database"
  ) +
  geom_hline(yintercept = 0) +
  theme_minimal() 

ggsave(here("figures", "01_ch_e0.png"), height = 20, width = 15, units = "cm", dpi = 300)
ggsave(here("figures", "01_ch_e0.svg"), height = 20, width = 15, units = "cm", dpi = 300)


```



# Mx at specific ages 

```{r}
dta_Mx %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  filter(age %in% c(0, 40, 80, 90)) %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y =  Mx, group = Country, colour = Country, linetype = Country)) +
  geom_line() +
  facet_grid(age ~ sex, scales = "free_y") +
  scale_y_log10() +
  theme_minimal() + 
  labs(
    x = "Year", y = "Probability of death in next 12 months (log scale)",
    title = "Probability of dying in next 12 months by age in years",
    caption = "Source: Human Mortality Database"
  )

ggsave(here("figures", "02_Mx_selected.png"), height = 20, width = 18, units = "cm", dpi = 300)
ggsave(here("figures", "02_Mx_selected.svg"), height = 20, width = 18, units = "cm", dpi = 300)



```
And swapping the groupings/facets 

```{r}
dta_Mx %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  filter(age %in% c(0, 40, 80, 90)) %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = year, y =  Mx, group = age, colour = as.factor(age), linetype = as.factor(age))) +
  geom_line() +
  facet_grid(Country ~ sex) +
  scale_y_log10() +
  theme_minimal() + 
  labs(
    x = "Year", y = "Probability of death in next 12 months (log scale)",
    title = "Probability of dying in next 12 months by age in years",
    caption = "Source: Human Mortality Database"
  )

ggsave(here("figures", "02a_Mx_selected.png"), height = 20, width = 18, units = "cm", dpi = 300)
ggsave(here("figures", "02a_Mx_selected.svg"), height = 20, width = 18, units = "cm", dpi = 300)



```

And indexed 

```{r}
dta_Mx %>% 
  left_join(country_labels) %>% 
  filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
  rename(Country = label) %>% 
  filter(age %in% c(0, 40, 80, 90)) %>% 
  filter(year >= 1975) %>% 
  filter(sex != "total") %>%
  group_by(Country, sex, age) %>% 
  mutate(indexed_Mx = 100 * Mx / Mx[year == 1975]) %>% 
  ungroup() %>% 
  ggplot(aes(x = year, y =  indexed_Mx, group = age, colour = as.factor(age), linetype = as.factor(age))) +
  geom_line() +
  facet_grid(Country ~ sex) +
  theme_minimal() + 
  labs(
    x = "Year", y = "Trends in probability of dying in next 12 months",
    title = "Age specifc mortality compared with 1975 (1975 = 100)",
    caption = "Source: Human Mortality Database"
  )

ggsave(here("figures", "02b_Mx_selected.png"), height = 20, width = 18, units = "cm", dpi = 300)
ggsave(here("figures", "02b_Mx_selected.svg"), height = 20, width = 18, units = "cm", dpi = 300)



```



Correlation between log improvement rates by country 

```{r}
corrstretch <- function(x){
  x %>% correlate() %>% stretch()
}

mx_corrs <- 
  dta_Mx %>% 
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
    filter(age %in% c(0, 40, 80, 90)) %>% 
    filter(year >= 1975) %>% 
    filter(sex != "total") %>% 
    mutate(lnmx = log(Mx)) %>%
    select(-Mx) %>% 
    select(Country, sex, age, year, lnmx) %>% 
    spread(age, lnmx) %>% 
    select(-year) %>% 
    group_by(sex, Country) %>% 
    nest() %>% 
    mutate(correlations = map(data, corrstretch)) %>% 
    select(Country, sex, correlations) %>% 
    unnest(correlations) 

mx_corrs  
```

Rearrange for table 

```{r}
corrs_female <- 
  mx_corrs %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "female") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "f_0", "f_40", "f_80", "f_90"))

corrs_male <- 
  mx_corrs %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "male") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "m_0", "m_40", "m_80", "m_90"))

corrs_wide <- full_join(corrs_female, corrs_male, by = c("Country", "age"))
corrs_wide
```

Now to display as a nice table 

```{r}
options(knitr.kable.NA = '')

corrs_wide %>% 
  kable(col.names = c("Country", "Age", "0", "40", "80", "90", "0", "40", "80", "90")) %>% 
  kable_styling(c("bordered")) %>% 
  add_header_above(c(" " = 2, "Female" = 4, "Male" = 4)) %>% 
  collapse_rows(columns = 1)
  
```


Now the same but split by period (1975-1990; 1991-last year)


```{r}
mx_corrs_1975_1990 <- 
  dta_Mx %>% 
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
    filter(age %in% c(0, 40, 80, 90)) %>% 
    filter(between(year, 1975, 1990)) %>% 
    filter(sex != "total") %>% 
    mutate(lnmx = log(Mx)) %>%
    select(-Mx) %>% 
    select(Country, sex, age, year, lnmx) %>% 
    spread(age, lnmx) %>% 
    select(-year) %>% 
    group_by(sex, Country) %>% 
    nest() %>% 
    mutate(correlations = map(data, corrstretch)) %>% 
    select(Country, sex, correlations) %>% 
    unnest(correlations) 

mx_corrs_1975_1990  

```

```{r}
corrs_female_1975_1990 <- 
  mx_corrs_1975_1990 %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "female") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "f_0", "f_40", "f_80", "f_90"))

corrs_male_1975_1990 <- 
  mx_corrs_1975_1990 %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "male") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "m_0", "m_40", "m_80", "m_90"))

corrs_wide_1975_1990 <- full_join(corrs_female_1975_1990, corrs_male_1975_1990, by = c("Country", "age"))
corrs_wide_1975_1990
```



Now to display as a nice table 

```{r}
options(knitr.kable.NA = '')

corrs_wide_1975_1990 %>% 
  kable(col.names = c("Country", "Age", "0", "40", "80", "90", "0", "40", "80", "90")) %>% 
  kable_styling(c("bordered")) %>% 
  add_header_above(c(" " = 2, "Female" = 4, "Male" = 4)) %>% 
  collapse_rows(columns = 1)

```


And for the latter period 


```{r}
mx_corrs_1991_last <- 
  dta_Mx %>% 
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
    filter(age %in% c(0, 40, 80, 90)) %>% 
    filter(year >= 1991) %>% 
    filter(sex != "total") %>% 
    mutate(lnmx = log(Mx)) %>%
    select(-Mx) %>% 
    select(Country, sex, age, year, lnmx) %>% 
    spread(age, lnmx) %>% 
    select(-year) %>% 
    group_by(sex, Country) %>% 
    nest() %>% 
    mutate(correlations = map(data, corrstretch)) %>% 
    select(Country, sex, correlations) %>% 
    unnest(correlations) 

mx_corrs_1991_last

```

```{r}
corrs_female_1991_last <- 
  mx_corrs_1991_last %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "female") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "f_0", "f_40", "f_80", "f_90"))

corrs_male_1991_last <- 
  mx_corrs_1991_last %>%
    mutate(r = round(r, 3)) %>% 
    spread(y, r) %>% 
  filter(sex == "male") %>%  
  ungroup() %>% select(-sex) %>% 
  set_names(nm = c("Country", "age", "m_0", "m_40", "m_80", "m_90"))

corrs_wide_1991_last <- full_join(corrs_female_1991_last, corrs_male_1991_last, by = c("Country", "age"))
corrs_wide_1991_last
```



Now to display as a nice table 

```{r}
options(knitr.kable.NA = '')

corrs_wide_1991_last %>% 
  kable(col.names = c("Country", "Age", "0", "40", "80", "90", "0", "40", "80", "90")) %>% 
  kable_styling(c("bordered")) %>% 
  add_header_above(c(" " = 2, "Female" = 4, "Male" = 4)) %>% 
  collapse_rows(columns = 1)

```



# Mx for all ages (heatmap)




```{r}
dta_trnd <- dta_Mx %>% 
  filter(sex != "total") %>%
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
  filter(year >= 1975)  %>% 
  filter(age <= 109) %>% 
  group_by(Country, sex, year, age) %>% 
  summarise(mean_Mx = mean(Mx, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(log_mean_Mx = log(mean_Mx, 10)) %>% 
  group_by(sex, Country) %>% 
  nest()


cors_df <- dta_trnd %>% 
  mutate(cors = map(
    data, 
    function(X) {
      X %>% 
        select(-mean_Mx) %>% 
        spread(age, log_mean_Mx) %>% 
        select(-year) %>% 
        cor()
      }
    )
  ) %>% 
  mutate(cor_df = map(
    cors,
    function(X){
      X %>% 
        as_tibble() %>% 
        mutate(from_age = rownames(X)) %>% 
        gather(key = "to_age", value = "value", -from_age) %>% 
        mutate(from_age = as.numeric(from_age), to_age = as.numeric(to_age))
      }
    )
  ) %>% 
  select(sex, Country, cor_df) %>% 
  unnest()


cors_df %>% 
  filter(from_age <= 100, to_age <= 100) %>% 
  rename(r = value) %>% 
  ggplot(aes(x = from_age, y = to_age, fill = r)) + 
  geom_tile() +
  scale_fill_distiller(palette = "RdBu", limits = c(-1,1), direction = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  coord_equal() + 
  facet_grid(Country ~ sex) +
  labs(x = "", y = "",
       title = "Correlation between % improvement\ntrends at specific ages on x and y axes")

ggsave(here("figures", "heatmap_allyears.png"), width = 18, height = 35, units = "cm", dpi = 300)
# ggsave(here("figures", "heatmap_allyears.svg"), width = 18, height = 35, units = "cm", dpi = 300)


```



Now the same for first 1975-1990, then 1991-last year 

```{r}
dta_trnd <- dta_Mx %>% 
  filter(sex != "total") %>%
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
  filter(between(year, 1975, 1990))  %>% 
  filter(age <= 109) %>% 
  group_by(Country, sex, year, age) %>% 
  summarise(mean_Mx = mean(Mx, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(log_mean_Mx = log(mean_Mx, 10)) %>% 
  group_by(sex, Country) %>% 
  nest()


cors_df <- dta_trnd %>% 
  mutate(cors = map(
    data, 
    function(X) {
      X %>% 
        select(-mean_Mx) %>% 
        spread(age, log_mean_Mx) %>% 
        select(-year) %>% 
        cor()
      }
    )
  ) %>% 
  mutate(cor_df = map(
    cors,
    function(X){
      X %>% 
        as_tibble() %>% 
        mutate(from_age = rownames(X)) %>% 
        gather(key = "to_age", value = "value", -from_age) %>% 
        mutate(from_age = as.numeric(from_age), to_age = as.numeric(to_age))
      }
    )
  ) %>% 
  select(sex, Country, cor_df) %>% 
  unnest()


cors_df %>% 
  filter(from_age <= 100, to_age <= 100) %>% 
  rename(r = value) %>% 
  ggplot(aes(x = from_age, y = to_age, fill = r)) + 
  geom_tile() +
  scale_fill_distiller(palette = "RdBu", limits = c(-1,1), direction = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  coord_equal() + 
  facet_grid(Country ~ sex) +
  labs(x = "", y = "",
       title = "Correlation between % improvement\ntrends at specific ages on x and y axes",
       subtitle = "1975-1990")

ggsave(here("figures", "heatmap_1975_1990.png"), width = 18, height = 35, units = "cm", dpi = 300)
# ggsave(here("figures", "heatmap_1975_1990.svg"), width = 18, height = 35, units = "cm", dpi = 300)


```


And now 1991 to last available year 

```{r}
dta_trnd <- dta_Mx %>% 
  filter(sex != "total") %>%
    left_join(country_labels) %>% 
    filter(label %in% c("USA", "Canada", "United Kingdom", "Japan", "France")) %>% 
    rename(Country = label) %>% 
  filter(year >= 1991)  %>% 
  filter(age <= 109) %>% 
  group_by(Country, sex, year, age) %>% 
  summarise(mean_Mx = mean(Mx, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(log_mean_Mx = log(mean_Mx, 10)) %>% 
  group_by(sex, Country) %>% 
  nest()


cors_df <- dta_trnd %>% 
  mutate(cors = map(
    data, 
    function(X) {
      X %>% 
        select(-mean_Mx) %>% 
        spread(age, log_mean_Mx) %>% 
        select(-year) %>% 
        cor()
      }
    )
  ) %>% 
  mutate(cor_df = map(
    cors,
    function(X){
      X %>% 
        as_tibble() %>% 
        mutate(from_age = rownames(X)) %>% 
        gather(key = "to_age", value = "value", -from_age) %>% 
        mutate(from_age = as.numeric(from_age), to_age = as.numeric(to_age))
      }
    )
  ) %>% 
  select(sex, Country, cor_df) %>% 
  unnest()


cors_df %>% 
  filter(from_age <= 100, to_age <= 100) %>% 
  rename(r = value) %>% 
  ggplot(aes(x = from_age, y = to_age, fill = r)) + 
  geom_tile() +
  scale_fill_distiller(palette = "RdBu", limits = c(-1,1), direction = 1) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  coord_equal() + 
  facet_grid(Country ~ sex) +
  labs(x = "", y = "",
       title = "Correlation between % improvement\ntrends at specific ages on x and y axes",
       subtitle = "1991 onwards")

ggsave(here("figures", "heatmap_1991_last.png"), width = 18, height = 35, units = "cm", dpi = 300)
# ggsave(here("figures", "heatmap_1991_last.svg"), width = 18, height = 35, units = "cm", dpi = 300)


```


# Workbooks 

It'll be good to be able to save the values from cors_df as a series of excel worksheets in workbooks. Let's look at the contents 

```{r}
cors_df %>% 
  ungroup() %>% 
  filter(sex == "female", Country == "Canada") %>% 
  select(from_age, to_age, value) %>% 
  spread(to_age, value)


```

Let's first produce a workbook with just the two periods visualised above 

```{r}
calc_and_grab_corrs <- function(DTA, this_country, this_sex, start_period, end_period, continuity_correction = 0){
  selected_data <- 
    DTA %>% 
      filter(sex == this_sex) %>% 
      filter(Country == this_country) %>% 
      filter(between(year, start_period, end_period)) %>% 
      filter(age <= 109) %>% 
      mutate(log_Mx = log(Mx + continuity_correction, 10)) %>% 
      select(year, age, log_Mx)  
    
  selected_data %>%
    spread(age, log_Mx) %>%
    select(-year) %>%
    cor()
}



dta_Mx_labelled_filtered <- 
  dta_Mx %>% 
  left_join(country_labels) %>% 
  select(Country = label, sex, year, age, Mx)

all_corrblocks <- 
  crossing(
    Country = c("France", "Japan", "United Kingdom", "USA", "Canada"),
    sex = c("female", "male"),
    start_period = c(1975, 1991),
    end_period = c(1990, 2019)
  ) %>% 
  filter(magrittr::or(
    start_period == 1975 & end_period %in% c(1990, 2019),
    start_period == 1991 & end_period == 2019    
    )
  ) %>% 
    mutate(corr_data = pmap(
      .l = list(this_country = Country, this_sex = sex, start_period = start_period, end_period = end_period), 
      .f = calc_and_grab_corrs, 
      DTA = dta_Mx_labelled_filtered
      )
    )


```

Write out these blocks

```{r}
add_format_worksheet <- function(wb, label, corr_data){
  wb %>% addWorksheet(sheetName = label, zoom = 25)
  wb %>% writeData(sheet = label, x = corr_data, rowNames = TRUE)
  wb %>% setColWidths(sheet = label, cols = 1:ncol(corr_data), widths = 3)
  wb %>% setRowHeights(sheet = label, rows = 1:(nrow(corr_data)), heights = 20)
  wb %>% conditionalFormatting(sheet = label, cols = 1:ncol(corr_data), rows = 1:(nrow(corr_data)),
                             type = "colorscale", style = c("red", "white", "blue"), rule = c(-1, 0, 1))
  
  NULL
}

wb <- openxlsx::createWorkbook()

all_corrblocks %>% 
  mutate(label = str_c(str_extract(Country, "^.{2}"), str_extract(sex, "^.{2}"), start_period, end_period, sep = "_")) %>% 
  mutate(tmp = walk2(label, corr_data, add_format_worksheet, wb = wb))

saveWorkbook(wb, file = here("tables", "blocks_both_main_periods.xlsx"), overwrite = TRUE)

```


Let's take a rolling series of 25 year periods, each five years apart, starting from 1950.

```{r}
calc_and_grab_corrs <- function(DTA, this_country, this_sex, start_period, window_length = 25, continuity_correction = 0){
  selected_data <- 
    DTA %>% 
      filter(sex == this_sex) %>% 
      filter(Country == this_country) %>% 
      filter(between(year, start_period, start_period + window_length)) %>% 
      filter(age <= 109) %>% 
      mutate(log_Mx = log(Mx + continuity_correction, 10)) %>% 
      select(year, age, log_Mx)  
    
  selected_data %>%
    spread(age, log_Mx) %>%
    select(-year) %>%
    cor()
}



dta_Mx_labelled_filtered <- 
  dta_Mx %>% 
  left_join(country_labels) %>% 
  select(Country = label, sex, year, age, Mx)

all_corrblocks <- 
  crossing(
    Country = c("France", "Japan", "United Kingdom", "USA", "Canada"),
    sex = c("female", "male"),
    start_period = seq(1950, 1995, by = 5)
  ) %>% 
    mutate(corr_data = pmap(
      .l = list(this_country = Country, this_sex = sex, start_period = start_period), 
      .f = calc_and_grab_corrs, 
      DTA = dta_Mx_labelled_filtered
      )
    )

```

Now to figure out how to write two of these in a neat way , before moving to others 

```{r, eval = FALSE}
all_corrblocks %>% 
  slice(1:2)

block_01 <- all_corrblocks %>% pull("corr_data") %>% pluck(1)
block_02 <- all_corrblocks %>% pull("corr_data") %>% pluck(2)

wb <- openxlsx::createWorkbook()

wb %>% addWorksheet(sheetName = "block01")
wb %>% writeData(sheet = "block01", x = block_01, startRow = 4, rowNames = TRUE)
wb %>% writeData(sheet = "block01", startRow = 1, x = all_corrblocks %>% slice(1) %>% select(-corr_data) %>%  mutate(end_period = start_period + 25))
wb %>% setColWidths(sheet = "block01", cols = 1:ncol(block_01), widths = 3)
wb %>% setRowHeights(sheet = "block01", rows = 4:(nrow(block_01)+4), heights = 20)
wb %>% conditionalFormatting(sheet = "block01", cols = 1:ncol(block_01), rows = 4:(nrow(block_01)+4),
                             type = "colorscale", style = c("red", "white", "blue"), rule = c(-1, 0, 1))

wb %>% addWorksheet(sheetName = "block02")
wb %>% writeData(sheet = "block02", x = block_02, startRow = 4, rowNames = TRUE)
wb %>% writeData(sheet = "block02", startRow = 1, x = all_corrblocks %>% slice(2) %>% select(-corr_data) %>% mutate(end_period = start_period + 25))
wb %>% setColWidths(sheet = "block02", cols = 1:ncol(block_02), widths = 3)
wb %>% setRowHeights(sheet = "block02", rows = 4:(nrow(block_02)+4), heights = 20)


saveWorkbook(wb, file = here("tables", "two_blocks.xlsx"), overwrite = TRUE)

```


Now for all blocks

```{r}
add_format_worksheet <- function(wb, label, corr_data){
  wb %>% addWorksheet(sheetName = label, zoom = 25)
  wb %>% writeData(sheet = label, x = corr_data, rowNames = TRUE)
  wb %>% setColWidths(sheet = label, cols = 1:ncol(corr_data), widths = 3)
  wb %>% setRowHeights(sheet = label, rows = 1:(nrow(corr_data)), heights = 20)
  wb %>% conditionalFormatting(sheet = label, cols = 1:ncol(corr_data), rows = 1:(nrow(corr_data)),
                             type = "colorscale", style = c("red", "white", "blue"), rule = c(-1, 0, 1))
  
  NULL
}

wb <- openxlsx::createWorkbook()

all_corrblocks %>% 
  mutate(label = str_c(str_extract(Country, "^.{2}"), str_extract(sex, "^.{2}"), start_period, sep = "_")) %>% 
  mutate(tmp = walk2(label, corr_data, add_format_worksheet, wb = wb))

saveWorkbook(wb, file = here("tables", "many_blocks.xlsx"), overwrite = TRUE)

```

# Linearity of trends 

A complementary analysis to that produced above is to investigate how (log)linear the trends have been at different ages. It's expected that trends in infancy will be largely (log)linear, along with those in retirement ages, whereas trends in working age will be least (log)linear, i.e. those age groups with the weakest correlations and strongest negative correlations with those at the majority of ages will also be those with where the (log)linearity of the trends is smallest. 

```{r}
get_rsq <- function(DTA, start_year, end_year){

  DTA %>% 
    filter(between(year, start_year, end_year)) %>% 
    mutate(logMx = log(Mx, 10)) %>% 
    lm(logMx ~ year, data = .) %>% 
    broom::glance() %>% 
    pull("r.squared")
}

all_rsq <- 
  dta_Mx_labelled_filtered %>% 
  filter(Country %in% c("France", "Japan", "United Kingdom", "USA", "Canada")) %>% 
    group_by(Country, sex, age) %>% 
    nest() %>% 
    crossing(
      start_year = c(1975, 1991),
      end_year   = c(1990, 2018)
    ) %>% 
    filter(!(end_year < start_year)) %>% 
    mutate(rsq = pmap_dbl(
      .l = list(DTA = data, start_year = start_year, end_year = end_year),
      .f = possibly(get_rsq, otherwise = NA_real_)
    ))

all_rsq

```


Now to visualise 

```{r}
all_rsq %>% 
  mutate(
    period = case_when(
      start_year == 1975 & end_year == 2018   ~ "1975 onwards",
      start_year == 1975 & end_year == 1990   ~ "1975-1990",
      start_year == 1991 & end_year == 2018   ~ "1991 onwards",
      TRUE                                    ~ NA_character_
    ) %>% factor(., levels = c("1975 onwards", "1975-1990", "1991 onwards"), ordered =TRUE)
  ) %>% 
  write_to_table(here("tables", "linearity_rsq.csv")) %>% 
  filter(sex != "total") %>% 
  ggplot(aes(x = age, y = rsq, group = period, linetype = period, colour = period)) + 
  geom_line() + 
  facet_grid(Country ~ sex) + 
  labs(
    x = "Age in years", 
    y = "Log-linearity (R-squared)",
    title = "Linearity in log mortality improvement rates"
  ) + 
  theme_minimal() +
  scale_colour_manual(values = c("black", "darkred", "darkblue")) +
  scale_linetype_manual(values = c("dashed", "dotdash", "solid"))

ggsave(here("figures", "log_linearity_by_age.png"), height = 20, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "log_linearity_by_age.svg"), height = 20, width = 20, units = "cm", dpi = 300)

```

From this figure the following features are apparent: 

* Overall, high linearity ($R^2$ values around 1) in trends in infancy and old age (from around 70 to 90 years) is apparent in all populations. 
* Over the late 80s and 90s the lienarity in trends reduces.
* In most populations the age at which linearity in trends starts to fall increases when comparing the earlier (1975-1990) to latter (1991 onwards) periods. This is evidence by comparing the fall off in linearity with age within a population of the dot-dashed red line with the solid blue line. It tends to start falling off at a higher age in the latter than the earlier period. 
* The USA is exceptional in that the trends are, from 1991, almost completely nonlinear in the late 20s for both sexes. For females this lack of linear improvement is also apparent in the early thirties, and in the mid 40s. 
* Overall the USA 

And let's do a less noisy version with just the full period 

```{r}
all_rsq %>% 
  mutate(
    period = case_when(
      start_year == 1975 & end_year == 2018   ~ "1975 onwards",
      start_year == 1975 & end_year == 1990   ~ "1975-1990",
      start_year == 1991 & end_year == 2018   ~ "1991 onwards",
      TRUE                                    ~ NA_character_
    ) %>% factor(., levels = c("1975 onwards", "1975-1990", "1991 onwards"), ordered =TRUE)
  ) %>% 
  filter(sex != "total") %>% 
  filter(period == "1975 onwards") %>% 
  ggplot(aes(x = age, y = rsq)) + 
  geom_line() + 
  facet_grid(Country ~ sex) + 
  labs(
    x = "Age in years", 
    y = "Log-linearity (R-squared)",
    title = "Linearity in log mortality improvement rates, 1975 onwards"
  ) + 
  theme_minimal() 

ggsave(here("figures", "log_linearity_by_age_1975onwards.png"), height = 20, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "log_linearity_by_age_1975onwards.svg"), height = 20, width = 20, units = "cm", dpi = 300)


```


And a third version overlaid 


```{r}
all_rsq %>% 
  mutate(
    period = case_when(
      start_year == 1975 & end_year == 2018   ~ "1975 onwards",
      start_year == 1975 & end_year == 1990   ~ "1975-1990",
      start_year == 1991 & end_year == 2018   ~ "1991 onwards",
      TRUE                                    ~ NA_character_
    ) %>% factor(., levels = c("1975 onwards", "1975-1990", "1991 onwards"), ordered =TRUE)
  ) %>% 
  filter(sex != "total") %>% 
  filter(period == "1975 onwards") %>% 
  ggplot(aes(x = age, y = rsq, group = Country, colour = Country)) + 
  geom_line() + 
  facet_grid(. ~ sex) + 
  labs(
    x = "Age in years", 
    y = "Log-linearity (R-squared)",
    title = "Linearity in log mortality improvement rates, 1975 onwards"
  ) + 
  theme_minimal() 

ggsave(here("figures", "log_linearity_by_age_1975onwards_overlaid.png"), height = 10, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "log_linearity_by_age_1975onwards_overlaid.svg"), height = 10, width = 20, units = "cm", dpi = 300)


```

```{r}
all_rsq %>% 
  mutate(
    period = case_when(
      start_year == 1975 & end_year == 2018   ~ "1975 onwards",
      start_year == 1975 & end_year == 1990   ~ "1975-1990",
      start_year == 1991 & end_year == 2018   ~ "1991 onwards",
      TRUE                                    ~ NA_character_
    ) %>% factor(., levels = c("1975 onwards", "1975-1990", "1991 onwards"), ordered =TRUE)
  ) %>% 
  filter(sex != "total") %>% 
  filter(period == "1975-1990") %>% 
  ggplot(aes(x = age, y = rsq, group = Country, colour = Country)) + 
  geom_line() + 
  facet_grid(. ~ sex) + 
  labs(
    x = "Age in years", 
    y = "Log-linearity (R-squared)",
    title = "Linearity in log mortality improvement rates, 1975-1990"
  ) + 
  theme_minimal() 

ggsave(here("figures", "log_linearity_by_age_1975_1990_overlaid.png"), height = 10, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "log_linearity_by_age_1975_1990_overlaid.svg"), height = 10, width = 20, units = "cm", dpi = 300)


```

From these it's clear that the USA has less linearity in trends in earlier working age, for both sexes, and that the USA is exceptional in this lack of linearity being observed in both males and females. The UK has a breakdown in linearity in males, and Canada, to a lesser extent, a breakdown in linearity in females. 

It is also apparant that the USA has a breakdown in linearity with old age from a younger age than France and Japan. 



```{r}
all_rsq %>% 
  mutate(
    period = case_when(
      start_year == 1975 & end_year == 2018   ~ "1975 onwards",
      start_year == 1975 & end_year == 1990   ~ "1975-1990",
      start_year == 1991 & end_year == 2018   ~ "1991 onwards",
      TRUE                                    ~ NA_character_
    ) %>% factor(., levels = c("1975 onwards", "1975-1990", "1991 onwards"), ordered =TRUE)
  ) %>% 
  filter(sex != "total") %>% 
  filter(period == "1991 onwards") %>% 
  ggplot(aes(x = age, y = rsq, group = Country, colour = Country)) + 
  geom_line() + 
  facet_grid(. ~ sex) + 
  labs(
    x = "Age in years", 
    y = "Log-linearity (R-squared)",
    title = "Linearity in log mortality improvement rates, 1991 onwards"
  ) + 
  theme_minimal() 

ggsave(here("figures", "log_linearity_by_age_1991_plus_overlaid.png"), height = 10, width = 20, units = "cm", dpi = 300)
ggsave(here("figures", "log_linearity_by_age_1991_plus_overlaid.svg"), height = 10, width = 20, units = "cm", dpi = 300)


```

For completeness perhaps % improvement (the beta?) might also be helpful. 



